### 2024-11-22 README.Rmd
update README
clean and remove modelr dependency
add useage to READEME

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

@@

# hatchR: Predicting Fish Developmental Phenology

@@

To come...

@@

* dplyr: For data manipulation.
* ggplot2: For vignette visualizations.
* lubridate: For working with dates.
* tibble: For creating tibbles as output for various functions.
* modelr: For modeling functions.
* rlang: For `.data` usage.
* stats: For various modeling functions (e.g., `confint()`).
* utils: For various utility functions.

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

@@

@@

Below we provide a brief example showing how to use `hatchR`. For a thorough introduction to the software, see our introductory vignette [linked here](https://bmait101.github.io/hatchR/articles/Introduction.html). For a list of all functions available in `hatchR`, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).

First we load `hatchR` and helper packages by running

```{r}
library(hatchR)
library(ggplot2)  # for additional plotting options
library(lubridate)  # for working with dates
```

The example `woody_island` dataset is included in `hatchR` and contains temperature data from Woody Island, Lake Iliamna, Alaska. It includes daily water temperature data from 1990-1992. We can use `plot_check_temp()` to visually check the data:

```{r}
p <- plot_check_temp(
  data = woody_island,
  dates = date, 
  temperature = temp_c
  )
p
```

Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year.

```{r}
p + 
  geom_rect(
    aes(
      xmin = ymd("1990-08-18"),  # spawn date
      xmax = ymd("1991-04-01"),  # approx phenology window end
      ymin = -10,  # lower bound
      ymax = 25),  # upper bound
    fill = "darkgreen",  # color rectangle dark green
    alpha = 0.01  # make rectangle semi-transparent
    ) 
```

Next, select hatch and emerge models we want to use to predict phenology:

```{r}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  dev.type = "hatch"
  )
```

Now we can predict phenology for the sockeye salmon using the `predict_phenology()` function:

```{r}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

We can check the predicted phenology by running:

```{r}
# see days to hatch
WI_hatch$days2done
# and development period
WI_hatch$dev.period
```

Finally, we can visualize the predicted phenology using `plot_phenology()`:

```{r}
plot_phenology(WI_hatch)
```

@@

* dplyr: For data manipulation.
* ggplot2: For vignette visualizations.
* lubridate: For working with dates.
* tibble: For creating tibbles as output for various functions.
* rlang: For `.data` usage.
* stats: For various modeling functions (e.g., `confint()`).
* utils: For various utility functions.
